---
title: "chromploid Package"
author: "Rosana Zenil-Ferguson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chromploid package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Chromploid is a package that allows for calculation of likelihoods  chromosome number or ploidy change models. This package allows user to choose among some fixed models but it also has the flexibility to extend to custom-user models.

Models included in the package are:

1. BiChroM: Chromosome number change associated with binary trait change (10 parameters)
2. PloidEvol: Ploidy change (6 parameters)
3. ChromEvol M3: Chromosome number change model M3 from Mayrose et al. (2010)^[A footnote here.]

## Installation

You'll need package ```devtools```. Chromploid is easily installed from github
```{r,error=FALSE, message=FALSE, warning=FALSE}
library("devtools")
devtools::install_github("roszenil/chromploid")
library("chromploid")
library("geiger") #chromploid depends on geiger so don't forget to load it
```

Chromploid package depends on libraries `ape`, `geiger`, and `expm` so make sure you are running those too.

## Negative log likelihood

The function ```neglolglikelihood()``` is possibly the most important and critical of the package. It calculates the negative log-likelihood of any discrete trait model for chromosome number or ploidy change. Minimizing negative log-likelihoods allows us to infer model parameters, that is, obtaining maximum likelihood estimates (MLEs), profile likelihoods, likelihood-confidence intervals, and likelihood ratio tests necessary to answer biological questions. I will describe how to use this important function  after introducing the models coded in chromploid. 

## BiChroM

Binary trait associated with chromosome number change model (BiChroM) is a continuous time Markov chain (CTMC) that allows for estimation and testing of chromosomal change rates correlated with a binary trait. The functions associated to BiChroM  model are:

* ```Q_bichrom()```: Creates the Q-matrix (instantaneous probability rate matrix) for later calculation of likelihood estimates
* ```Q_unichrom()```: Creates the marginal Q-matrix necessary for calculation of univariate profile likelihoods
* ```Q_bibichrom()```: Creates the marginal Q-matrix necessary for calculation of bivariate profile likelihoods
* ```Q_reducedbichrom()```: Creates the marginal Q-matrix necessary for calculation of likelihood ratio tests.
* ```bichrom_dataset()```: Transforms a dataset of chromosome and binary traits to input in the functions above

### Full inference for BiChroM model

1. Description of parameters in BiChroM

  * $\lambda_0, \lambda_1$: Rate of gain in one chromosome for taxa with binary trait 0, or binary trait 1 respectively.
  * $\mu_0, \mu_1$: Rate of loss in one chromosome for taxa with binary trait 0, or binary trait 1 respectively.
  * $\rho_0, \rho_1$: Rate of chromosome doubling for taxa with binary trait 0, or binary trait 1 respectively.
  * $q_{01},q_{10}$: Rate of transition between binary state values, from 0 to 1 or vice versa.
  * $\epsilon_0, \epsilon_1$: Ancillary parameters  of chromosomal changes. Rate of chromosome number changes after ```size``` has been determined.
  
2. BiChroM simulation and estimation
  * Choose a set of fixed parameter values.
For example $(\lambda_0,\lambda_1, \mu_0,\mu_1,\rho_0,\rho_1,q_{01},q_{10},\epsilon_0,\epsilon_1)=$
$(0.12, 0.001, 0.25, 0.002,0.036, 0.006, 0.04,0.02, 1.792317852, 1.57e-14)$
  
  * Calculate BiChroM Q-matrix
  
```{r, message=FALSE, warning=FALSE}
log.params<-log(c(0.12, 0.001, 0.25, 0.002,0.036, 0.006, 0.04,0.02, 1.792317852, 1.57E-14))
N<- 50
mymatrix<-Q_bichrom(log.theta=log.params, size=N)
```
  
  * Simulate a tree with 500 taxa and BiChroM values 
  Assuming that the root of the tree has a value 56 means that the root is has 5 haploid chromosomes and its binary type is 1. 
  
```{r}
mytree<- sim.bdtree(b=0.055, stop="taxa", n=500, seed=1) # Seed was fixed to obtain always the same tree
plot(mytree,type="fan",cex=0.2, no.margin=TRUE)
set.seed(1) #Seed was fixed to obtain always the same sample
mysample<- sim.char2(phy=mytree, par=mymatrix,model="discrete", root=56)
```
  
  * Calculate the likelihood of the tree, assuming that the root has a uniform distribution for chromosome number and binary state ```pi.0=NULL``` (other options available). But first a very common error
  
```{r, eval=FALSE}
nllike<-negloglikelihood(log.par=log.params, phy=mytree, tip.values=mysample,pi.0=NULL,Q.FUN=Q_bichrom, size=50)
> Error in tip.values[i, charnum + 1] : incorrect number of dimensions
```
  The error above indicates that you don't have in the first column the species names and in the second column the sample for BiChrom. An easy way to fix it is converting your sample into a data frame with first column having taxa names and second column having the sample needed in BiChroM
```{r}
mysample<-data.frame(taxa=rownames(mysample),sample=as.numeric(mysample))
head(mysample)
```
  Then re-run the argument to calculate the negative log-likelihood ```nllike```
```{r}
nllike<-negloglikelihood(log.par=log.params, phy=mytree, tip.values=mysample,pi.0=NULL,Q.FUN=Q_bichrom, size=50)
nllike
```

3. BiChrom optimization
If you have a tree and a sample and you need to find the maximum likelihood estimates of the parameters determined by BiChroM you can use any optimizer of your choice. I suggest using package ```nloptr``` but ```optim``` from ```stats``` package works fine as long as you manipulate correctly the options. Here's an example using ```nloptr```.

```{r, eval=FALSE}
library(nloptr)
x.0<- log(c(0.12, 0.001, 0.25, 0.002,0.036, 0.006, 0.04,0.02, 1.792317852, 1.57e-14)) #value where the optimization algorithm is going to start. A vector of 10 values for the parameters in log scale
results<-rep(0,11) #vector useful to save optimization results
my.options<-list("algorithm"= "NLOPT_LN_SBPLX","ftol_rel"=1e-08,"print_level"=1,"maxtime"=170000000, "maxeval"=1000) # Options used in nloptr, for more information ?nloptr. This is a subplex algorithm, with high tolerance to fine a very precise optimum.
mle<-nloptr(x0=x.0,eval_f=negloglikelihood,opts=my.options, phy=mytree, tip.values=mysample,pi.0=NULL,Q.FUN=Q_bichrom, list(size=50)) 
print(mle) 
results[1:10]<-exp(mle$solution) # Maximum likelihood estimates in original scale 
results[11]<-mle$objective # Negative log-likelihood value at the MLEs (minimum)
```

